\Large Inferential Analysis of the Supreme Court Citation Network

### 2018-04-09 Tex/paper.tex
fixed some typos

$$h_{covariate}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

The structure of this model is very similar to a conventional (T)ERGM setup. The main difference is that  $\mathcal{C}_t$ excludes adjacency matrices in which there are loops that include edges sent at different time points. Loops can exist in the Supreme Court citation network, but only among cases decided in the same term. Furthermore, a case $i$ can only cite a case $j$ if $j$ entered the network before or at the same time as $i$ did. The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with almost $10,000$ vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.

@@

We include two sets of dummy variables that account for the issue areas of cases. In one set of dummy variable---sender intercepts---the variable $X_{ij}$ indicates the issue area of the sending case ($i$). In the set of receiver intercepts $X_{ij}$ indicates the issue are of the receiver case. Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} find that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} states which number represents which issue area.

@@

We model the way in which citations to a case depend upon the age of a case. For this we use a second-order polynomial in which one covariate $X_{ij}$ is defined as the age of case $j$ at the time that case $i$ is decided, and another term in which $X_{ij}$ is the squared age of case $j$ at the time that case $i$ is decided. We include these covariates because \citet{black2013citation} find that the number of citations to a Supreme Court case over time depends significantly on the age of the case, characterized by a sharp drop off and leveling out with age.

\citet{benjamin2012standing} study the propensity for cases to be overruled and cited in other negative ways. They find that cases with majority coalitions that are large and ideologically broad are less likely to be cited negatively. In our data we do not differentiate between negative and positive citations, but since the overwhelming majority of citations are positive, we hypothesize that the effects they found will be reversed in our analysis. We include one covariate in which $X_{ij}$ is the number of justices in the majority coalition for case $j$. We also include another covariate in which $X_{ij}$ is the absolute difference between the maximum and minimum ideal points of the justices in the majority coalition for case $j$. We expect both of these covariates to have positive effects.

@@

 \section{Empirical Analysis}
Our three data sources for this study include the Supreme Court Database (SCDB) \citep{spaeth2014supreme}, Martin-Quinn scores \citep{martin2002dynamic}, and Supreme Court citation data provided by the CourtListener Free Law Project \citep{CourtListener}. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015)\footnote{There were 145 cases that were listed in the SCDB but could not be matched to a case in the CourtListener data. We decided to exclude these 145 cases from our model.}. The Supreme Court citation network from $1937 - 2015$ consists of $9,945$ cases. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. In order to only account for positive citations, we exclude $315$ citations that caused the cited case to be overruled. As a result, the network has a total of $111,986$ ties. The number of triangles in the network is $250,717$ and the number of mutual ties is $56$. The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in Figure \ref{degree_dist}. The maximum indegree is $230$ and the maximum outdegree is $162$.

@@

The effects of the exogenous covariates included in the c-TERGM (panels (e)--(k) in Figure \ref{fig:coeftrends_exo}) exhibit some surprising dynamics. The effects of absolute difference in Martin-Quinn score, ideological breadth of the recipient case, and majority size of the recipient case all flip signs throughout the time period under study. This may be an artifact of the simple linear trends we have estimated, but is interesting nonetheless. The attitudinal model holds that ideological preferences drive decision-making on the US Supreme Court. Ideological selection of precedents would be evidenced by a negative effect of the absolute difference in MQ scores in the c-TERGM. We find this effect is negative and statistically significant only after 1975. The absolute difference in MQ scores variable is on the scale of approximately 0---3. At its most ideological point (i.e. 2001), the probability that case $i$ cites case $j$ decreases from approximately 0.024 to 0.018---by one third---if the ideological distance between the majority coalitions in the two cases increases by 3. The effects of the majority size and ideological breadth of case $j$ are similar in both scale and trend. It is not until later in the time series that they both exhibit their expected positive effects. The effect of same issue is positive and statistically significant, and substantial for the time period under study. The probability that $i$ cites $j$ increases by approximately three-fold over the naive probability of citation if $i$ and $j$ have the same legal issue area.

>>>>

$$h_{covariate}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}X_{\leq t}^{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.\\
\indent The structure of this model is very similar to a conventional (T)ERGM setup. The main difference is that  $\mathcal{C}_t$ excludes adjacency matrices in which there are loops that include edges sent at different time points. Loops can exist in the Supreme Court citation network, but only among cases decided in the same term. Furthermore, a case $i$ can only cite a case $j$ if $j$ entered the network before or at the same time as $i$ did. \\
\indent The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with almost $10,000$ vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.

@@

We include two sets of dummy variables that account for the issue areas of cases. In one set of dummy variable---sender intercepts---the variable $X^{ij}$ indicates the issue area of the sending case ($i$). In the set of receiver intercepts $X^{ij}$ indicates the issue of the receiver case. Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} finds that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} states the different issue areas.

@@

We model the way in which citations to a case depend upon the age of a case. For this we use a second-order polynomial in which one covariate $X^{ij}$ is defined as the age of case $j$ at the time that case $i$ is decided, and another term in which $X^{ij}$ is the squared age of case $j$ at the time that case $i$ is decided. We include these covariates because \citet{black2013citation} find that the number of citations to a Supreme Court case over time depends significantly on the age of the case, characterized by a sharp drop off and leveling out with age.

\citet{benjamin2012standing} study the propensity for cases to be overruled and cited in other negative ways. They find that cases with majority coalitions that are large and ideologically broad are less likely to be cited negatively. In our data we do not differentiate between negative and positive citations, but since the overwhelming majority of citations are positive, we hypothesize that the effects they found will be reversed in our analysis. We include one covariate in which $X^{ij}$ is the number of justices in the majority coalition for case $j$. We also include another covariate in which $X^{ij}$ is the absolute difference between the maximum and minimum ideal points of the justices in the majority coalition for case $j$. We expect both of these covariates to have positive effects.

@@

 \section{Empirical Analysis}
Our three data sources for this study include the Supreme Court Database (SCDB) \citep{spaeth2014supreme}, Martin-Quinn scores \citep{martin2002dynamic}, and Supreme Court citation data provided by the CourtListener Free Law Project \citep{CourtListener}. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015)\footnote{There were 145 cases that were listed in the SCDB but could not be matched to a case in the CourtListener data. We decided to exclude these 145 cases from our model.}. The Supreme Court citation network from $1937 - 2015$ consists of $9,945$ cases and the breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. In order to only account for positive citations, we exclude $315$ citations that caused the cited case to be overruled. As a result, the network has a total of $111,986$ ties. The number of triangles in the network is $250,717$ and the number of mutual ties is $56$. The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in Figure \ref{degree_dist}. The maximum indegree is $230$ and the maximum outdegree is $162$.

@@

The effects of the exogenous covariates included in the c-TERGM (panels (e)--(k) in Figure \ref{fig:coeftrends_exo}) exhibit some surprising dynamics. The effects of absolute difference in Martin-Quinn score, and majority size of the recipient case all flip signs throughout the time period under study. This may be an artifact of the simple linear trends we have estimated, but is interesting nonetheless. The attitudinal model holds that ideological preferences drive decision-making on the US Supreme Court. Ideological selection of precedents would be evidenced by a negative effect of the absolute difference in MQ scores in the c-TERGM. We find this effect is negative and statistically significant only after 1975. The absolute difference in MQ scores variable is on the scale of approximately 0---3. At its most ideological point (i.e. 2001), the probability that case $i$ cites case $j$ decreases from approximately 0.024 to 0.018---by one third---if the ideological distance between the majority coalitions in the two cases increases by 3. The effects of the majority size and ideological breadth of case $j$ are similar in both scale and trend. It is not until later in the time series that they both exhibit their expected positive effects. The effect of same issue is positive and statistically significant, and substantial for the time period under study. The probability that $i$ cites $j$ increases by approximately three-fold over the naive probability of citation if $i$ and $j$ have the same legal issue area.
